Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-01-06 paper/tidy-missing-data-paper.Rmd
tidy up section 2

There are many ways each value can go missing. The information in $R$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), Missing at random, and missing not at random. **MCAR** is where values being missing have no association with observed or unobserved data, that is $Pr(missing | observed, unobserved) = Pr(missing)$. Essentially, the probability of an observation being missing is unrelated to anything else.  Although this is a convenient scenario, it is not actually possible to confirm, as it relies on statements on data unobserved. **MAR** refers to cases where the missingness only depends on the data observed, and not data unobserved, that is,$Pr(missing | observed, unobserved) = Pr(missing | observed)$. Some structure or dependence between the missing values and observed values is allowed, provided that this can be explained by the data observed. Finally, **MNAR** refers to the missingness being related to values unobserved: $Pr(missing | observed, unobserved)$. This assumes that conditioning on all available observed data, the data goes missing due to some phenomena unobserved. This presents a challenge in analysis, as it is difficult to verify this state, and also implies bias in analysis due to the unobserved phenomena.

@@

\pkg{VIM} [@VIM] efficiently implements well-used imputation methods k-nearest neighbor, regression, hot-deck, and iterative robust model-based imputation. This diversity of approaches allow the package to deal semi-continuous, continuous, count, and categorical data. VIM identifies imputed cases by adding an indicator variable with a suffix `_imp`. So `Var1` would have a sibling indicator column, `Var1_imp`, with TRUE or FALSE values to indicate imputed value. \pkg{VIM} also has a variety of visualization methods, discussed in section \@ref(exploration). The \pkg{simputation} package includes imputation methods for linear regression, decision trees, K nearest neighbors, hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977]. A key feature of its design is that it returns a regular dataframe. This approach contrasts from other imputation software, which may provide custom classes of data or additional objects, which require additional work to prepare for subsequent analysis. Providing a dataframe with imputed values included in the data reduces the friction of working with other tools, but comes at the cost identifying imputed values. Other software for imputation includes \pkg{Hmisc} [@Hmisc], which provide predictive mean matching, \pkg{imputeTS} [@imputeTS], which provide time series imputation methods, and \pkg{missMDA} [@missMDA], which imputes data using principal components analysis.

@@

Multiple imputation is often regarded as best practice for imputing values [@Schafer2002], as long as appropriate caution is taken [@Sterne2009]. Popular and robust methods for multiple imputation include the \pkg{mice}, \pkg{Amelia}, and \pkg{mi} packages [@mice; @amelia; @mi]. The \pkg{mice} package implements the method of chained equations, using a variable-wise algorithm to calculate the posterior distribution of parameters, from which uses to generate imputed values. The suggested workflow in \pkg{mice} revolves around imputing data, returning completed data, and fitting a model and pooling the results. The \pkg{Amelia} package [@amelia] assumes data are multivariate normal, and samples from the posterior using the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm [@Honaker2010] and allows for incorporation of information on the values in a prior. The \pkg{mi} package [@mi] also uses Bayesian models for imputation, providing better handling of semi-continuous values, and data with structural or perfect correlation. A collection of analysis models are also provided in \pkg{mi}, which work with the specially multiply imputed data, including linear models, generalized linear models, and their corresponding Bayesian components. This approach promotes fluid workflow, with a similar penalty to tidying up model output, which is still messy.  Another software for performing multiple imputation is \pkg{norm} [@norm; @schafer-norm], which uses methods from the NORM software from @schafer-norm, such as multiple imputations using EM for multivariate normal data. The \pkg{norm} package does not provide a framework for tracking missing values, but instead provides tools for making inference from multiple imputation. Each of these multiple imputation methods provide practical methods, but do not have consistent interfaces across their methods or implementation, which makes then inherently messy, as their output is always different. A consistent data structure that tracks missing values by default would make the data outputs tidy, streamlining subsequent analysis.

@@

The primary focus of most missing data packages is exploring imputed values, not
on exploring relationships in missing values, and identifying possible patterns.
Texts that do cover exploration of missing data have the same problem as with modelling: the input is tidy, but the output is messy and does not work with other tools [@VanBuuren2012]; this is inefficient. Methods for exploring missing values are primarily covered in literature on interactive graphics [@Swayne1998; @Unwin1996; @Cook2007], and are picked up again in a discussion of a graphical user interface [@Cheng2015].

Missing values in a dataset can be represented as a binary matrix where 0 is present and 1 is missing [@Rubin1976; @Schafer2002]. This is often referred to as a matrix $R$, which can be used to assess missing data dependence. This idea of a matrix of missing values has also been used in interactive graphics, dubbed a "shadow matrix", to link missing and imputed values to the data, facilitating their display in graphics [@Swayne1998]. This work also focuses heavily on multivariate numeric data. This is an idea upon which this new work builds.

The approach of @Unwin1996 focussed on multivariate categorical data, where missingness is explicitly added as an additional category. MANET also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot. The approach of @Swayne1998 in the software XGobi, further developed in ggobi [@Cook2007], focussed more on multivariate quantitative data. Missingness is incorporated into plots in ggobi by setting them to be 10% below the minimum value.

MissingDataGUI provides a user interface for exploring missing data structure both numerically and visually. Using a GUI to explore missing data may make it easier to glean valuable insight into important structures for missingness. However, these insights are typically not captured or recorded, so it is difficult to incorporate them into reproducible analyses. This distracts and breaks analysis workflow, inviting mistakes.

VIM (visualizing and Imputing Missing Data) provides visualization methods that identify observed, imputed, and missing values. These include spinograms, spinoplots, missingness matrices, plotting missingness in the margins of other plots, and other summaries. These help explore missingness, however these visualizations do not map variables to graphical aesthetics, creating friction moving across workflows, making them difficult to extend to new circumstances. Additionally, data used to create the visualizations cannot be accessed, posing a barrier to further explorations.

\pkg{ggplot2} incorporates missingness into visualizations only when mapping a discrete variable to a graph aesthetic. This has some limitations. For example, for data with school grades and test scores, a boxplot visualization could have school grade mapped to the x axis, and test score to the y axis (See Figure \@ref(fig:gg-box-na)). If there are missings in a continuous variable like test score, \pkg{ggplot2} omits the missing values and prints a warning message. However, if a discrete variable like school year has missing values, an NA category is created for school year, and the scores are placed into this.

Figure \@ref(fig:gg-box-na), shows four scenarios of missingness for test scores across school year - the first contains no missingness, the second contains missingness for school year, the third contains missingness in scores, and the fourth contains missingness in both.

@@

cowplot::plot_grid(p1,
          p2,
          p3,
          p4,
          nrow = 2,
          labels = LETTERS[1:4])
```

>>>>

There are many ways each value can go missing, we adopt the notation used in @VanBuuren2012. The information in $R$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), Missing at random (MAR), and missing not at random (MNAR). The distribution of missing values in $r_{ij}$ can depend on the entire dataset, which is represented as $Y = (Y_{obs}, Y_{miss})$. This relationship can be defined by the _missing data model_, of which $/psi$ contains the model parameters, and the missing data model can written as $Pr(r_{ij} | Y_{obs}, Y_{miss}, \psi)$ - the probability of missingness is conditional on the data observed, data missing, and some probability parameter of missingness. We can then use this to help precisely define categories of missing values.

**MCAR** is where values being missing have no association with observed or unobserved data, that is $Pr(R = 1 | Y_{obs}, Y_{miss}) = Pr(R = 1 | \psi)$. Essentially, the probability of an observation being missing is unrelated to anything else, only the parameters $\psi$, which is the overall probability of missingness. Although this is a convenient scenario, it is not actually possible to confirm, or clearly distinguish from MAR, as it relies on statements on data unobserved.

**MAR** refers to cases where the missingness only depends on the data observed, and not data missing, that is,$Pr(R = 1 | Y_{obs}, Y_{miss}, \psi) = Pr(R = 1 | Y_{obs}, \psi)$. Some structure or dependence between the missing values and observed values is allowed, provided that this can be explained by the data observed, and some overall probability of missingness.

Finally, **MNAR** refers to the missingness being related to values observed, and un unobserved; if this does not simplify  $Pr(R = 1 | Y_{obs}, Y_{miss}, \psi)$. This assumes that conditioning on all available observed data, the data goes missing due to some phenomena unobserved, including the structure of the missing data itself. This presents a challenge in analysis, as it is difficult to verify this state, and also implies bias in analysis due to the unobserved phenomena.

@@

\pkg{VIM} [@VIM] efficiently implements well-used imputation methods k-nearest neighbor, regression, hot-deck, and iterative robust model-based imputation. This diversity of approaches allow the package to deal semi-continuous, continuous, count, and categorical data. VIM identifies imputed cases by adding an indicator variable with a suffix `_imp`. So `Var1` would have a sibling indicator column, `Var1_imp`, with TRUE or FALSE values to indicate imputed value. \pkg{VIM} also has a variety of visualization methods, discussed in section \@ref(exploration). The \pkg{simputation} package provides an interface into popular imputation methods for linear regression, decision trees, K nearest neighbors, hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977]. Other software for imputation includes \pkg{Hmisc} [@Hmisc], which provide predictive mean matching, \pkg{imputeTS} [@imputeTS], which provide time series imputation methods, and \pkg{missMDA} [@missMDA], which imputes data using principal components analysis.

A key feature of the design of \pkg{simputation} is that it returns a regular dataframe. This is different from other imputation packages, which return custom classes of data or additional objects. These create extra work when preparing for subsequent analysis. Providing a dataframe with imputed values included in the data reduces the friction of working with other tools, but comes at the cost identifying imputed values.

@@

Multiple imputation is often regarded as best practice for imputing values [@Schafer2002], as long as appropriate caution is taken [@Sterne2009]. Popular and robust methods for multiple imputation include the \pkg{mice}, \pkg{Amelia}, and \pkg{mi} packages [@mice; @amelia; @mi]. The \pkg{mice} package implements the method of chained equations, using a variable-wise algorithm to calculate the posterior distribution of parameters to generate imputed values. The workflow in \pkg{mice} revolves around imputing data, returning completed data, and fitting a model and pooling the results.

The \pkg{Amelia} package [@amelia] assumes data are multivariate normal, and samples from the posterior using the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm [@Honaker2010] and allows for incorporation of information on the values in a prior.

The \pkg{mi} package [@mi] also uses Bayesian models for imputation, providing better handling of semi-continuous values, and data with structural or perfect correlation. A collection of analysis models are also provided in \pkg{mi}, which work with the specially multiply imputed data, including linear models, generalized linear models, and their corresponding Bayesian components. This approach promotes fluid workflow, with a similar penalty to tidying up model output, which is still messy.

Another software for performing multiple imputation is \pkg{norm} [@norm; @schafer-norm], which uses methods from the NORM software from @schafer-norm, such as multiple imputations using EM for multivariate normal data. The \pkg{norm} package does not provide a framework for tracking missing values, but instead provides tools for making inference from multiple imputation.

Each of these multiple imputation methods provide practical methods, but all have different output structures, and do not have consistent interfaces in their implementation. This makes then inherently messy and challenging to integrate into an analysis pipeline. An ideal approach would have consistent data structures that work with other analysis tools, and track missing values by default. This would make the data outputs tidy, streamlining subsequent analysis.

@@

The primary focus of most missing data packages is making inferences, and exploring imputed values, not
on exploring relationships in missing values, and identifying possible patterns.
Texts that do cover exploration of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools [@VanBuuren2012]; this is inefficient. Methods for exploring missing values are primarily covered in literature on interactive graphics [@Swayne1998; @Unwin1996; @Cook2007], and are picked up again in a discussion of a graphical user interface [@Cheng2015].

The missingness matrix $R$ can be used to assess missing data dependence, and has  been used in interactive graphics, dubbed a "shadow matrix", to link missing and imputed values to the data, facilitating their display in graphics [@Swayne1998]. This work also focuses heavily on multivariate numeric data. This is an idea upon which this new work builds.

The approach of the MANET software in @Unwin1996 focussed on multivariate categorical data, where missingness is explicitly added as an additional category. MANET also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot. The MANET software is no longer maintained and cannot be installed. The approach of @Swayne1998 in the software XGobi, further developed in ggobi [@Cook2007], focussed more on multivariate quantitative data. Missingness is incorporated into plots in ggobi by setting them to be 10% below the minimum value.

MissingDataGUI provides a user interface for exploring missing data structure both numerically and visually. Using a Graphical User Interface (GUI) to explore missing data may make it easier to glean valuable insight into important structures for missingness. However, this comes as a trade off with GUIs - the insights are typically not captured or recorded, so it is difficult to incorporate them into reproducible analyses. This distracts and breaks analysis workflow, inviting mistakes.

VIM (Visualizing and Imputing Missing Data) provides visualization methods that identify observed, imputed, and missing values. These include spinograms, spinoplots, missingness matrices, plotting missingness in the margins of other plots, and other summaries. These help explore missingness, however these visualizations do not map variables to graphical aesthetics, creating friction moving across workflows, making them difficult to extend to new circumstances. Additionally, data used to create the visualizations cannot be accessed, posing a barrier to further explorations.

\pkg{ggplot2} incorporates missingness into visualizations only when mapping a discrete variable to a graph aesthetic. This has some limitations. For example, for data with school grades and test scores, a boxplot visualization could have school grade mapped to the x axis, and test score to the y axis (See Figure \@ref(fig:gg-box-na)). If there are missings in a continuous variable like test score, \pkg{ggplot2} omits the missing values and prints a warning message. However, if a discrete variable like school year has missing values, an NA category is created for school year, and the scores are placed into this (Figure \@ref(fig:gg-box-na)).

@@

cowplot::plot_grid(p1,
          p2,
          p3,
          p4,
          nrow = 1,
          labels = LETTERS[1:4])
```
